\fontfamily{phv}\selectfont{Causal Discovery methods for Effective Connectivity in Human Brains}

### 2014-01-17 docs/paper/paper final.tex
wat TODOs afgewerkt en data acquisition verplaatst

\begin{abstract}
Many different approaches have been used to infer brain connectivity.
But few general causal discovery methods have yet been used to find effective connectivity.
In this article, we describe how the PC algorithm is used on functional brain-data.
This approach results in connectivity matrices which clearly show some well-known brain connections.
Some disadvantages of the PC algorithm for this use are discussed, and an attempt is made to decrease these disadvantages by making adjustments to the algorithm.
Causal relations have only be found with very low certainty. 
% TODO: first state what you try to do, then how you did it, then what it brought you
%In this work we applied PC algorithm on resting-state data of healthy human subjects.
%Additionally we propose a variant of PC algorithm that poses weaker assumptions on the model, better fitting the underlying model.
%We find that detecting structure homological areas in the brain works better with PC algorithm than most conventional diffusion weighted imaging techniques.
%In the area of directionality and causality, more research is needed.
\end{abstract}

@@

Ramsey et al. lists several problems in inferring brain connectivity.
%- Indirect measurements
One of these problems is that many interesting brain-data is measured indirectly; a measured variable is an effect of the cause we want to know.
This makes inference hard in practical models: many latent variables may be present, influencing this cause-effect relation.
%- Distinct but overlapping ROIs
%- Varying Haemodynamic response (HRF)
Related is the problem of haemodynamic response fluctuations (HRF).
In functional MRI-measurements, the delay between brain activity and the indirect measurement of that activity is unknown. %TODO moest MRI nou helemaal uitgeschreven of niet? Ik vind zelf van niet.
Any form of causal inference based on sequence in time might therefore be incorrect; A cause may be measured after its effects, if it has a larger delay.
%- Modelling causal structure across individuals (varying responses to same stimuli)
Also, brain-data is not always consistent, both over time and across different persons. This makes it hard to get significant results.
%Find shortcomings of PC and improve it (EMS-PC)

@@

\paragraph{Shortcomings}
Although PC is proven correct, the version provided above as standard PC is not.
It is also not complete.
There is additional the number of possible graphs in the returned Markov class increasing exponentially with the amount of variables in the provided data, thus not always providing ample information.
Also, standard PC is dependent on the order in which these variables are presented, as this is the same order in which edges between pairs of these variables are considered for removal\cite[p.88]{spirtes2000}.

In practice, the lack of full correctness and completeness are of little concern as these involve only very specific and little occurring connectivity patterns\cite[p.127-130]{spirtes2000}.
% TODO: Paragraph seems incomplete and some sentences are weird.

@@

@@

\begin{figure}[h!]
  \centering
  \includegraphics[height=0.28\textwidth]{images/new/struct_subj1-crop} % TODO remove colorbar, why is it even there
  \includegraphics[height=0.28\textwidth]{images/new/struct_subj2-crop}
  \includegraphics[height=0.28\textwidth]{images/new/struct_subj3_colorbar-crop} % one colorbar at the top right of the figure
  \includegraphics[height=0.28\textwidth]{images/new/struct_subj4-crop}
  \includegraphics[height=0.28\textwidth]{images/new/struct_subj5-crop}
  \includegraphics[height=0.28\textwidth]{images/new/struct_subj6-crop}
  \caption{The structure of all six subjects as found with the PC algorithm. The two diagonal lines represent connections between the two hemispheres.}
\label{fig:struct_subjects}
\end{figure}

@@

As can be seen in figures~\ref{assym_subjects} and~\ref{assym_avg}, the certainties of one-directional arrows are very low compared to the certainties of the bi-directional arrows. Also, the certainties of one-directional arrows are somewhat higher in EMS-PC than in standard PC.
For EMS-PC, the highest certainties found for one-directional edges per subject are in the range of 0.45 to 0.9, whereas the avarage across subjects has a maximum value of 0.3.
The few one-directional arrows found in idividual subjects therefore do not seem to be consistent across different subjects.

@@

\appendix
\section{Data acquisition}\label{sec:data}
%Subject acquisition and instruction
%Used equipment and specs
%Experimental setup
%Scanner settings
%Preprocessing
%- Software
%- Scripts
%Resulting data specs
%TODO niet helemaal happy, hoe zeg je dit zonder steeds "as described by Hinne et al." te zeggen? "This research" kan ook naar ons eigen werk verwijzen en is verwarrend, het moet duidelijk zijn dat dit alles niet door ons gedaan is.
We performed our analysis on the same data as treated by Hinne et al. in the aforementioned Bayesian structural analysis \cite{hinne2013, hinne2013structfunc}.
This article describes how resting-state functional data of six subjects was measured at the Donders Centre for Cognitive Neuroimaging at the Radboud University.
This functional data is divided in 116 nodes and 1029 time-series per subject.
Additionally, diffusion-weighted images of these subjects have been obtained.
Based on these DWI-measurements, a structural connectivity matrix has been inferred with a technique described by Hinne et al.
We used the six resulting resting-state functional time series data and connectivity matrices from the Bayesian analysis as pre-processed by Hinne et al.

No additional pre-processing has been done in our research in order to apply the PC algorithm on this data.

>>>>

\begin{abstract}
Many different approaches have been used to infer brain connectivity.
But few general causal discovery methods have yet been used to find effective connectivity.
In this article, we describe how the PC algorithm is used on functional brain-data.
This approach results in connectivity matrices which clearly show some well-known brain connections.
Some disadvantages of the PC algorithm for this use are discussed, and an attempt is made to decrease these disadvantages by making adjustments to the algorithm.
Causal relations have only be found with very low certainty. 
%In this work we applied PC algorithm on resting-state data of healthy human subjects.
%Additionally we propose a variant of PC algorithm that poses weaker assumptions on the model, better fitting the underlying model.
%We find that detecting structure homological areas in the brain works better with PC algorithm than most conventional diffusion weighted imaging techniques.
%In the area of directionality and causality, more research is needed.
\end{abstract}

@@

Ramsey et al. lists several problems in inferring brain connectivity.
%- Indirect measurements
One of these problems is that many interesting brain-data is measured indirectly; a measured variable is an effect of the cause we want to know.
This makes inference hard in practical models: many latent variables may be present, influencing this cause-effect relation.
%- Distinct but overlapping ROIs
%- Varying Haemodynamic response (HRF)
Related is the problem of haemodynamic response fluctuations (HRF).
In functional MRI-measurements, the delay between brain activity and the indirect measurement of that activity is unknown.
Any form of causal inference based on sequence in time might therefore be incorrect; A cause may be measured after its effects, if it has a larger delay.
%- Modelling causal structure across individuals (varying responses to same stimuli)
Also, brain-data is not always consistent, both over time and across different persons. This makes it hard to get significant results.
%Find shortcomings of PC and improve it (EMS-PC)

@@

\paragraph{Shortcomings}
Although PC is proven correct, the version provided above as standard PC is not.
It is also not complete.
The number of possible graphs in the returned Markov class increases exponentially with the amount of variables in the provided data, thus the network resulting from PC might not be sufficient to describe the underlying structure.
Also, standard PC is dependent on the order in which these variables are presented, as this is the same order in which edges between pairs of these variables are considered for removal\cite[p.88]{spirtes2000}.

In practice, the lack of full correctness and completeness are of little concern as these involve only very specific and little occurring connectivity patterns\cite[p.127-130]{spirtes2000}.

@@

\section{Data acquisition}\label{sec:data}
%Subject acquisition and instruction
%Used equipment and specs
%Experimental setup
%Scanner settings
%Preprocessing
%- Software
%- Scripts
%Resulting data specs
We performed our analysis on the same data as treated by Hinne et al. in the aforementioned Bayesian structural analysis \cite{hinne2013, hinne2013structfunc}.
This article describes how resting-state functional data of six subjects was measured at the Donders Centre for Cognitive Neuroimaging at the Radboud University.
This functional data is divided in 116 nodes and 1029 time-series per subject.
Additionally, diffusion-weighted images of these subjects have been obtained.
Based on these DWI-measurements, a structural connectivity matrix has been inferred with a technique described by Hinne et al.
We used the six resulting resting-state functional time series data and connectivity matrices from this Bayesian analysis.

All measurements and pre-processing as described in this section has been done during previous research. Noadditional pre-processing was needed in our research in order to apply the PC algorithm on this data.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[height=0.28\textwidth]{images/new/struct_subj1-crop}
  \includegraphics[height=0.28\textwidth]{images/new/struct_subj2-crop}
  \includegraphics[height=0.28\textwidth]{images/new/struct_subj3_colorbar-crop} % one colorbar at the top right of the figure
  \includegraphics[height=0.28\textwidth]{images/new/struct_subj4-crop}
  \includegraphics[height=0.28\textwidth]{images/new/struct_subj5-crop}
  \includegraphics[height=0.28\textwidth]{images/new/struct_subj6-crop}
  \caption{The structure of all six subjects as found with the PC algorithm. The two diagonal lines represent connections between the two hemispheres.}
\label{fig:struct_subjects}
\end{figure}

@@

As can be seen in figures~\ref{fig:assym_subjects} and~\ref{fig:assym_avg}, the certainties of one-directional arrows are very low compared to the certainties of the bi-directional arrows. Also, the certainties of one-directional arrows are somewhat higher in EMS-PC than in standard PC.
For EMS-PC, the highest certainties found for one-directional edges per subject are in the range of 0.45 to 0.9, whereas the avarage across subjects has a maximum value of 0.3.
The few one-directional arrows found in idividual subjects therefore do not seem to be consistent across different subjects.

@@
